Viscous Erosion in Porous Media

### 2018-03-20 docs/methods/erosion_methods.tex
NM

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress grows as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / a$, in which drag is inversely proportional to body-scale $a$ (e.g.~for a sphere, $a$ is taken to be the radius). This law must be modified in two dimensions, though. The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $\ell$, the drag is proportional to
$\mu \umax \ell/ \log(\ell/a)$, where $a$ is a length scale of the cross-section and $a \ll \ell$. This relationship implies that the stress-per-unit-span on a 2D cross-section can be estimated as
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{a \log(\ell/a)} \qquad \text{for } a \ll \ell \, .
\end{align}
In interpreting our 2D simulations, we will treat the span, $\ell$, as
fixed. Recall that $L$ denotes the total perimeter of the 2D cross-section. Since $L$ and $a$ each represent a characteristic scale of the cross-section, they can often be interchanged in scaling analysis.

@@

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align}
\dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, ,
\end{align}
Thus, the rate of area reduction is given by the product of the mean shear stress $\mean{\atau}$ and the perimeter $L$, which is an {\em exact} relation. Inserting scaling law~\eqref{eqn:stresslaw} gives the estimate
\begin{align}
\dot{A} \sim \frac{ - \mu \umax}{\log(\ell/a)} \qquad \text{as } a \to 0 \, . 
\end{align}
Noting that $A \sim a^2$ allows one to solve the ODE, giving
\begin{align}
\label{area_predict}
\frac{A}{A_0} \left( 1 - c_1 \log{\frac{A}{A_0}} \right) \sim \frac{t_f - t}{t_f} \quad \text{as } t \to t_f \, .
\end{align}
Here, $t_f$ is the time of vanishing, $A_0$ is the body's initial area, and $c_1$ is a free constant. The above is an {\em implicit} relationship for how area scales down in time. We can use this prediction to test our numerical simulation. In Fig.~\ref{area_drag}(a) we compare the area measured in our numerical simulation (black curve) against the above scaling law (red dotted curve) with $c_1 = 0.24$ a fit parameter. The two agree remarkable well throughout the entire simulation, thus co-validating the simulation and asymptotic results.

	The drag exerted by the eroding body offers a second check on the simulation. Figure~\ref{area_drag}(b) shows the horizontal drag, $F_D$, calculated in the simulations (black curve) using Eq.~\eqref{eqn:drag}. In the context of slender-body theory, $F_D$ should be interpreted as drag-per-unit-span. As seen in Fig.~\ref{area_drag}(b), the drag decreases gradually at first and then abruptly tends to zero in the very final moments. This behavior, though perhaps surprising, is consistent with the fact that drag vanishes {\em logarithmically} with body-size in 2D Stokes flow. Equation \eqref{eqn:drag} also allows us to calculate the separate contributions of pressure and viscous drag, and we show these with the blue and green curves in Fig.~\ref{area_drag}(b). Initially, pressure and viscous drag are comparable, with the circular geometry having a slightly higher level of pressure drag. As the body erodes and assumes a more slender form, though, the pressure drag decreases much more rapidly, leaving viscous drag to dominate. This behavior is associated with the tendency towards a drag-minimizing profile, which will be discussed in Section \ref{LimitingShape}.

As a check on these results, we can use slender-body theory to estimate the drag-per-unit-span
\begin{align}
\label{dragscaling}
\frac{F_D}{4 \pi} \sim \frac{ \mu U}{\log(\ell/a)}	\qquad \text{for } a \ll \ell \, .
\end{align}
For the case of a {\em circular} cylinder, the normalization factor of $4 \pi$ produces an {\em exact} relationship (under the slender-body assumptions), with $a$ taken to be the radius. For other shapes the formula will be approximate and we define $a = \sqrt{A/\pi}$ as the characteristic lengthscale (since it generalizes the case of a circle). In Fig.~\ref{area_drag}(b) we show this scaling prediction (red dotted curve), where $\ell = 2.3 a_0$ is a fit parameter and $a_0 = 0.2$ is the radius of the initially circular body. The scaling law and measured drag agree closely as the body vanishes.

>>>>

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress grows as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, in which stress, $\tau \sim \mu \umax / a$, scales inversely proportional to body-scale $a$ (e.g.~for a sphere, $a$ is the radius). This law must be modified in two dimensions, though. The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $\ell$, the drag is proportional to
$\mu \umax \ell/ \log(\ell/a)$, where $a$ is a length scale of the cross-section and $a \ll \ell$. This relationship implies that the stress-per-unit-span on a 2D cross-section can be estimated as
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{a \log(\ell/a)} \qquad \text{for } a \ll \ell \, .
\end{align}
In interpreting our 2D simulations, we will treat the span, $\ell$, as
fixed. Recall that $L$ denotes the total perimeter of the 2D cross-section. Since $L$ and $a$ each represent a characteristic scale of the cross-section, they can often be interchanged in scaling analysis.

@@

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align}
\dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, ,
\end{align}
Thus, the rate of area reduction is given by the product of the mean shear stress $\mean{\atau}$ and the perimeter $L$, which is an {\em exact} relation. Inserting scaling law~\eqref{eqn:stresslaw} gives the estimate
\begin{align}
\dot{A} \sim \frac{ - \mu \umax}{\log(\ell/a)} \qquad \text{as } a \to 0 \, . 
\end{align}
Noting that $A \sim a^2$ allows one to solve the ODE, giving
\begin{align}
\label{area_predict}
\frac{A}{A_0} \left( 1 - c_1 \log{\frac{A}{A_0}} \right) \sim \frac{t_f - t}{t_f} \quad \text{as } t \to t_f \, .
\end{align}
Here, $t_f$ is the time of vanishing, $A_0$ is the body's initial area, and $c_1$ is a free constant. The above is an {\em implicit} relationship for how area scales down in time, and we can use this prediction to test our numerical simulation. In Fig.~\ref{area_drag}(a) we compare the area measured in our numerical simulation (black curve) against the above scaling law (red dotted curve) with $c_1 = 0.24$ a fit parameter. The two agree remarkable well throughout the entire simulation, thus co-validating the simulation and asymptotic results.

	Another important quantity that can be used to test our simulations is the drag exerted by the body as it erodes. Figure~\ref{area_drag}(b) shows the horizontal drag, $F_D$, calculated in the simulations (black curve) using Eq.~\eqref{eqn:drag}. In the context of slender-body theory, $F_D$ should be interpreted as drag-per-unit-span. As seen in Fig.~\ref{area_drag}(b), the drag decreases gradually at first and then abruptly tends to zero in the very final moments. This behavior, though perhaps surprising, is consistent with the fact that drag vanishes {\em logarithmically} with body-size in 2D Stokes flow. Equation \eqref{eqn:drag} also allows us to calculate the separate contributions of pressure and viscous drag, and we show these with the blue and green curves in Fig.~\ref{area_drag}(b). Initially, pressure and viscous drag are comparable, with the circular geometry having a slightly higher level of pressure drag. As the body erodes and assumes a more slender form, though, the pressure drag decreases much more rapidly, leaving viscous drag to dominate. This behavior is associated with the tendency towards a drag-minimizing profile, which will be discussed in Section \ref{LimitingShape}.

As a check on these results, we can use slender-body theory to estimate the drag-per-unit-span
\begin{align}
\label{dragscaling}
\frac{F_D}{4 \pi} \sim \frac{ \mu U}{\log(\ell/a)}	\qquad \text{for } a \ll \ell \, .
\end{align}
For the case of a {\em circular} cylinder, the normalization factor of $4 \pi$ produces an {\em exact} relationship (under the slender-body assumptions), where $a$ is taken to be the radius. For other shapes the formula will be approximate and we define $a = \sqrt{A/\pi}$ as the characteristic lengthscale since it generalizes the case of a circle. In Fig.~\ref{area_drag}(b) we show this scaling prediction (red dotted curve), where $\ell = 2.3 a_0$ is a fit parameter and $a_0 = 0.2$ is the radius of the initially circular body. The scaling law and measured drag agree closely as $t \to t_f$, thus confirming that our simulation accurately captures the slender-body limit as the shape disappears.
